White Wine Exploratory Data Analysis by Irene Florez
========================================================

# Introduction 

This project is a quick exploratory analysis project using R. 
The aim of the project is to use R to explore the relationships between 
data features.

We will be using a wine data set available at 
http://www3.dsi.uminho.pt/pcortez/wine/.

'White Wine Quality' is a tidy dataset which contains 4,898 white wines
with 11 variables quantifying the chemical properties of each wine. 
At least 3 wine experts rated the quality of each wine, providing 
a rating between 0 (very bad) and 10 (very excellent). 

We will use univariate, bivariate, and multivariate analyses to explore
the relationships between the data features and to tease out the quality 
rating.  You can read the final summary and reflection at the end of this
document. 

## Sources
Created by: Paulo Cortez (Univ. Minho), Antonio Cerdeira, Fernando Almeida, Telmo Matos and Jose Reis (CVRVV) @ 2009

* https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt
* http://www.wineskills.co.uk/winemaking/winemaking-knowledge-base/chemical-composition
* https://waterhouse.ucdavis.edu/whats-in-wine/red-wine-composition

## setup document
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=9, fig.height=5, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

## load packages
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages

library(ggplot2)
library(dplyr)
library(knitr)
library(reshape2)
library(GGally)
library(gridExtra)
library(memisc)

```

## load the data
```{r echo=FALSE, load_data}
# Load the Data
df = read.csv('wineQualityWhites.csv', header = T, row.names = 1)
```

## data variables
```{r echo=FALSE, names}
# review names of variables
dim(df)
names(df)
```

Dataset dimensions:
There are 12 variables & a total of 4898 observations

```{r echo = FALSE, structure}
#data structure
str(df)
```

Dataset structure:
All of the data observations are num or int, there are not factor data types

## data dimensions

```{r echo=FALSE, summarize}
#dimensions
head(df, 5)
summary(df)
```

Dataset features:
We will review quality, alcohol, sulphates, density, and sugar. 

# Univariate Plots Section

```{r, echo=FALSE}
#plot.hist function
plot.hist <- function(xvariable, df, bins) 
  {ggplot(df, aes_string(x = xvariable)) + 
   geom_histogram(bins = bins)
}
```

## quality
```{r echo=FALSE, quality_levels}
#check levels 
levels(df$quality)
```

```{r echo=FALSE, quality_summary}
#summary of quality
summary(df$quality)
```

Quality values are distributed between 3-9
The mean is 5.88 and the median is 6.00.

```{r echo=FALSE, quality_barchart}
#quality bar chart
ggplot(aes(x = quality), data = df) + 
  geom_bar() + 
  scale_x_continuous(breaks = seq(2, 10, 1))
```

Quality has a roughly normal bell shaped curve distribution
The largest frequency scored 6(44.88%) and
a small number of wines scored 9(0.1%).

## alcohol

```{r, echo=FALSE, alcohol1}
#alchohol summary
summary(df$alcohol)
```

Alcohol values range from 8-14.20
The avarage alcohol value is 10.51

```{r, echo=FALSE, alcohol2}
#alcohol histogram
hist(df$alcohol)
```

The largest group by frequency, has a 9-9.5 alcohol count.
There is a high concentration of wines between 8.5-12.5 alcohol count. 

## sulphates
```{r, echo=FALSE, sulphates1}
summary(df$sulphates)
```

Sulphate values range from .22 to 1.08
The average sulphate value is .48

```{r echo=FALSE, sulphate3}
#sulphate histogram
ggplot(aes(x = sulphates), data = df) + geom_histogram(binwidth = 0.01)              
```

The majority of wines have between .3 and .6 sulphate content. 

## density
```{r, echo=FALSE, density1}
#density summary
summary(df$density)
```

White wine density falls between .98 - 1.03, with a mean of .99

```{r echo=FALSE, density2}
#density histogram
ggplot(aes(x = density), data = df) + geom_histogram(binwidth = 0.01)              
```

This histogram is shifted to the left, which means density variable has at least one outlier. 


```{r, echo=FALSE, density3}
#density histogram views and scatter boxplot
s0 = plot.hist('density', df, 30)
s1 =  ggplot(df, aes(y = density, x = 1)) + 
  geom_jitter(aes(x = 1, y = density), alpha = 0.15) +
  geom_boxplot(color = 'blue', alpha = 1/4) + coord_flip()
s2 = plot.hist('density', df, 75) 
grid.arrange(s0, s1, s2)
```

The density content is closely grouped together for the majority of wines. 


## residual_sugar
```{r, echo=FALSE, residual.sugar1}
summary(df$residual.sugar)
```

Most wines have 1.7-9.9 residual sugar content. 
There are big outliers as far as residual sugar. 

```{r echo=FALSE, residual_sugar2}
#residual.sugar histogram
ggplot(aes(x = residual.sugar), data = df) +
  geom_histogram(bins = 30, color = "white")
```

As with density, the histogram of the amount of residual suger 
is shifted to the left. This means residual sugar may contain outliers.
Let's exclude the top 1% of residual sugar values... 

```{r, echo=FALSE, residual_sugar4}
#residual.sugar histogram, excluding the top 1%
ggplot(aes(x = residual.sugar),
       data = subset(df, residual.sugar < quantile(df$residual.sugar,0.99))) +
  geom_histogram(bins = 30, color = "white")
```
 
 We see that wines have 1-19 content of residual sugar. 
 
```{r, echo=FALSE, residual_sugar5}
#residual sugar scatter boxplot

s1 =  ggplot(df, aes(y = residual.sugar, x = 1)) + 
  geom_jitter(aes(x = 1, y = residual.sugar), alpha = 0.15) +
  geom_boxplot(color = 'blue', alpha = 1/4) + coord_flip()

grid.arrange(s1)
```
 
 Here we can see that the higher end of the outliers far 
 outpace the majority of the scores. 

# Bivariate Plots Section

```{r echo=FALSE, bivariate_plots1}
#correlation grid looking at all variables
ggcorr(df, nbreaks = 4, palette = "RdGy", label = TRUE, label_size = 3, 
       label_color = "white")
```

Density and alcohol have the strongest relationship
Alchohol has a negative relationship with density.

## alcohol & quality 

```{r echo=FALSE, bivariate_alcohol1} 
#correlation of alcohol and quality
with(df, cor.test(alcohol, quality, method = 'pearson'))

```

Alchohol has a positive relationship (.43) with quality. 

```{r echo=FALSE, bivariate_alchohol3}
#quality/alcohol scatter boxplot
ggplot(aes(x= factor(quality), y = alcohol ), data = df) +
geom_jitter( alpha = .1)  +
  geom_boxplot( alpha = .3,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  geom_smooth(se = FALSE, method = "lm", aes(group=1),
              color = "darkgreen")
```


```{r echo=FALSE, bivariate_alcohol5}
#plot the alcohol vs quality
ggplot(aes(x = alcohol, y = quality), data = df) + 
  geom_jitter(alpha = 0.1,color = 'orange') + 
  geom_smooth()
```

Now let's remove the outliers

```{r echo=FALSE, bivariate_alcohol6}
#plot the alcohol vs quality
#trimming big outliers
ggplot(aes(x = alcohol, y = quality), data = df) + 
  geom_jitter(alpha = 0.1, color = 'orange') + 
  xlim(quantile(df$alcohol, 0.01), quantile(df$alcohol,0.99)) +
  geom_smooth()
```

Generally, higher quality wines will have higher alchohol content. 


## sulphates & quality


```{r echo=FALSE, bivariate_sulphate1}
#plot sulphates vs quality
ggplot(aes(x= factor(quality), y = sulphates ), data = df) +
geom_jitter( alpha = .1)  +
  geom_boxplot( alpha = .3,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```

Generally, as quality increases, sulphate content increases...

Now let's remove outliers

```{r echo=FALSE, bivariate_sulphate2}
#boxplot sulphates vs quality
#removing big outliers
ggplot(aes(x = quality, y = sulphates), data = df) + 
  geom_boxplot(aes(group = quality)) + 
  ylim(quantile(df$sulphates, 0.01), quantile(df$sulphates, 0.99))
```

Generally, as quality increases, sulphate content increases...
... until quality scores reach 7.5   

## alcohol & sulphates

```{r echo=FALSE, bivariate_sulphate_alcohol1}
#pearson method of correlation analysis
with(subset(df, df$alcohol>11), cor.test(sulphates, 
                                         alcohol, method = 'pearson'))
```

```{r echo=FALSE, bivariate_sulphate_alcohol2}
#plot sulphates and alcohol
ggplot(aes(alcohol, sulphates), data = df) +
  geom_point(position = 'jitter', alpha= 0.1)
```
Alcohol and sulphates have a weak relationship. 

## density, & sugar

```{r echo=FALSE, bivariate_sugar1}
#plot density and residual.sugar
ggplot(aes(x = density, y =residual.sugar), data = df) + 
    geom_point(position = 'jitter', alpha= 0.1)
  #geom_boxplot(aes(group = quality))
```

Density and residual sugar contents are both low count variables. 
Generally, the higher the density, the higher the residual sugar content. 

```{r echo=FALSE}
#Change the quality variable to an ordered factor
df$quality.int <- df$quality
df$quality <- factor(df$quality, ordered = TRUE)
str(df$quality)

```

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

## alcohol, sulphates, & quality

```{r echo=FALSE, multivariate_alcohol_sulphate}
#plot alcohol, sulphates, and quality
ggplot(aes(x = alcohol, 
           y = sulphates, color = factor(quality)), 
       data = df) + 
  ylim(quantile(df$sulphates, 0.01), quantile(df$sulphates, 0.99)) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))
```

Alchohol has a positive relationship with quality. 


## alcohol, density, & quality

```{r echo=FALSE, multivariate_alcohol_density}
#plot alcohol, density, and quality
ggplot(aes(x = alcohol, 
           y = density, color = factor(quality)), 
       data = df) + 
  ylim(quantile(df$density, 0.01), quantile(df$density, 0.99)) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))
```

Generally, higher quality wines will have higher alchohol content. 
Alchohol has a negative relationship with density.


```{r echo=FALSE, multivariate_alcohol_density_quality}
# scatterplot - density and alcohol, by quality levels
ggplot(aes(density, alcohol, color = quality), data = df) +
  geom_point()  +
  scale_color_brewer(palette = "RdYlBu", name="Quality",
                     direction=-1) 
  #http://colorbrewer2.org/#type=sequential&scheme=BuPu&n=7

```



Findings: 
Alchohol has a positive relationship (.43) with quality. 
Generally, higher quality wines will have higher alchohol content. 
Alchohol has a negative relationship with density.
Wine density has a negative relationship with the quality.
wines with a higher score tend to have lower density.

Process:
We can see that there are outliers in the density variable. 
We will remove them for the final plots portion. 
We will also add a darker background, for contrast. 
This more clearly highlights the difference between good and bad wines (because a neutral color is used for the OK wines) and the levels in quality are highlighted by the gradation in color.


Because quality is ordinal, we set ggplot() to use sequential or divergent color encoding (this is optimal because it gives a sense of gradation to the different levels in the data) - and not qualitative color encoding (which is for general discrete variables).

Where sequential color encoding is used for pure ordinal discrete data and divergent color encoding is used if the data is both ordinal and follows a diverging scale (think "Good, Ok, Bad" - which can be viewed as appropriate for this dataset). For example, for the plot above:
(where "RdYlBu" is a specific divergent color scheme, the name option changes the legend title and direction=-1 changes the order of the colors) 







# Final Plots and Summary

## plot one
```{r echo=FALSE, final_correlation}
#chart correlation of all variables
ggcorr(df, geom = "blank", label = TRUE, hjust = 0.75) +
  geom_point(size = 10, aes(color = coefficient > 0, 
  alpha = abs(coefficient) > 0.3)) +
  scale_alpha_manual(values = c("TRUE" = 0.25, "FALSE" = 0)) +
  guides(color = FALSE, alpha = FALSE) 
  ggtitle('Relationships & Correlations') 
```
Alcohol and density have the strongest relationshop; these two are neagatively 
related. Density and alcohol are the strongest determinants of quality. 

The amount of residual sugar has a weak relationship wih the quality of wine.


## plot two
```{r echo=FALSE, final_alcohol_q_scatter_boxplot}
#plot quality and alcohol
ggplot(aes(x= factor(quality), y = alcohol ), data = df) +
geom_jitter( alpha = .1)  +
  geom_boxplot( alpha = .3,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  ggtitle('Relationship & distribution of alcohol and quality') +
  labs(x="Alcohol (% by Volume)", y = "quality (score between 0 and 10)")
```

There's a high concentration of 5-7 quality wines... 
And these largely have alchohol levels between 8.5-13.5 



## plot three
```{r echo=FALSE, final_scatterplot}
# scatterplot - density and alcohol, by quality levels
ggplot(aes(density, 
           alcohol, 
           color = quality), 
       data = subset(df, 
                     density < quantile(df$density, 0.99)
                     )
       ) +
  geom_point()  +
  scale_color_brewer(palette = "RdYlBu", name="Quality",
                     direction=-1) +
  theme(panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = "darkgrey")) +
  xlab(expression(Density~(g/cm^{3}))) +
  ylab("Alcohol (% by Volume)") 
  ggtitle('Relationship & distribution of alcohol, density, and quality') 
#http://colorbrewer2.org/#type=sequential&scheme=BuPu&n=7

      
```

Findings:
Alchohol has a positive relationship (.43) with quality.
Generally, higher quality wines will have higher alchohol content. 
Alchohol has a negative relationship with density.
Density of wine has a negative relationship with the qaulity of wine.

Process:
We removed the outliers in the density variable and added a darker background, for contrast. 
This more clearly highlights the difference between good and bad wines (because a neutral color is used for the OK wines) and the levels in quality are highlighted by the gradation in color.

Because quality is ordinal, we set ggplot() to use sequential or divergent color encoding (this is optimal because it gives a sense of gradation to the different levels in the data) - and not qualitative color encoding (which is for general discrete variables).

Where sequential color encoding is used for pure ordinal discrete data and divergent color encoding is used if the data is both ordinal and follows a diverging scale (think "Good, Ok, Bad" - which can be viewed as appropriate for this dataset). For example, for the plot above:
(where "RdYlBu" is a specific divergent color scheme, the name option changes the legend title and direction=-1 changes the order of the colors) 



# Reflection

EDA results: We reviewed six features in this data set: quality, alcohol, density, sulphates, 
and residual sugar. 
1) Density is the best predictor of quality. Higher quality wines tend to have 
lower density and higher alcohol content.
2) Alchohol has a positive relationship (.43) with quality. 
Generally, higher quality wines will have higher alchohol content. 
3) Sulphates and alchohol content generally do not have a strong relationship. 
4) The amount of residual sugar has a weak relationship wih the quality of wine.
Overall, higher quality wines tend to have low residual sugar content. 
Generally, as density increases, residual sugar will also increase


Process reflection: Here we used plotting (scatterplots, histograms, 
boxplots, and line graphs). 

In the future, I would like to do a correlation matrix using all of the variables. 
I would also like to review ratios such as quality / alcohol. Secondly, I would 
like to do a deeper dive into alcohol and density.  Lastly, I would like to create 
a model which with observations can predict wine quality with some degree of confidence. 
And overall, I would like more practice with transparency, jitter, smoothing, and 
limiting axes. 
