White Wine Exploratory Data Analysis by Irene Florez
========================================================

# Introduction 

This project is a quick exploratory analysis project using R. 
The aim of the project is to use R to explore the relationships between 
data features.

We will be using a wine data set available at 
http://www3.dsi.uminho.pt/pcortez/wine/.

'White Wine Quality' is a tidy dataset which contains 4,898 white wines
with 11 variables quantifying the chemical properties of each wine. 
At least 3 wine experts rated the quality of each wine, providing 
a rating between 0 (very bad) and 10 (very excellent). 

We will use univariate, bivariate, and multivariate analyses to explore
the relationships between the data features and to tease out the quality 
rating.  You can read the final summary and reflection at the end of this
document. 

## setup document
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

## load packages
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages

library(ggplot2)
library(dplyr)
library(knitr)
library(reshape2)
library(GGally)
library(gridExtra)
library(memisc)



```

## Load the data
```{r echo=FALSE, load_data}
# Load the Data
df = read.csv('wineQualityWhites.csv', header = T, row.names = 1)
```

# Univariate Plots Section

## data variables
```{r echo=FALSE, names}
# review names of variables
dim(df)
names(df)
```

# there are 12 variables & a total of 4898 observations

## data dimensions

```{r echo=FALSE, summarize}
#dimensions
head(df, 5)
summary(df)
```


```{r echo = FALSE, structure}
#data structure
str(df)
```

#all of the data observations are num or int, there are not factor data types

## features review: quality
```{r echo=FALSE, qualityLevels}
#check levels 
levels(df$quality)
```

```{r echo=FALSE, qualitySummary}
#5 number summary of quality
summary(df$quality)
```

# quality is a roughly normal bell shaped curve distribution
# quality values are distributed between 3-9

```{r echo=FALSE, quality_histogram}
ggplot(data = df, mapping = aes(x=quality)) + 
  geom_histogram(binwidth=1)
```
```{r echo=FALSE}
ggplot(aes(x = quality), data = df) + 
  geom_bar() + 
  scale_x_continuous(breaks = seq(2, 10, 1))
```

## features review: alcohol

```{r, echo=FALSE }
summary(df$alcohol)
```

```{r, echo=FALSE }
fah = plot.hist('alcohol', df, 100)  
fah1 = plot.hist('alcohol', df, 100) + 
  scale_x_continuous(limits = c(4, 12), breaks = seq(4, 12, 1))
grid.arrange(fah, fah1)
```

```{r echo=FALSE, alcohol1}
grid.arrange(ggplot(aes(x = alcohol), data = df) + 
  geom_histogram() ,
  ggplot(aes( x = 1, y = density ), data = df ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ), ncol = 2)
  
```

```{r echo=FALSE, alcohol2}
ggplot(aes(x = alcohol), data = df) + 
  geom_histogram(binwidth = 0.1) + 
  scale_x_continuous(breaks = seq(8, 14,1))
  
```

## features review: sulphates
```{r, echo=FALSE, sulphates1}
summary(df$sulphates)
```

```{r, echo=FALSE, sulphates2}
s0 = plot.hist('sulphates', df, 30)
s1 =  ggplot(df, aes(y = sulphates, x = 1)) + 
  geom_jitter(aes(x = 1, y = sulphates), alpha = 0.15) +
  geom_boxplot(color = 'blue', alpha = 1/4) + coord_flip()
s2 = plot.hist('sulphates', df, 75) 
grid.arrange(s0, s1, s2)
```


```{r echo=FALSE, sulphate3}
grid.arrange(ggplot(aes(x = sulphates), data = df) + 
  geom_histogram(),
ggplot(aes( x = 1, y = density ), data = df ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ), ncol = 2)
```

```{r echo=FALSE, sulphate4}
ggplot(aes(x = sulphates), data = df) + geom_histogram(binwidth = 0.01)              
```

# Univariate Analysis

# Bivariate Plots Section

## alcohol & quality 

```{r echo=FALSE, bivariate_alcohol1} 
with(df, cor.test(alcohol, quality, method = 'pearson'))

```

```{r echo=FALSE, bivariate_alcohol2}
#scatterplot of quality vs alcohol
ggplot(aes(quality, alcohol), data = df) +
  geom_jitter(alpha = 0.1)
```

```{r echo=FALSE, bivariate_alchohol3}
ggplot(aes(x= factor(quality), y = alcohol ), data = df) +
geom_jitter( alpha = .1)  +
  geom_boxplot( alpha = .3,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```

```{r echo=FALSE, bivariate_alcohol4}
ggplot(aes(x = quality, y = alcohol), data = df) + 
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, bivariate_alcohol5}
ggplot(aes(x = alcohol, y = quality), data = df) + 
  geom_jitter(alpha = 0.1,color = 'orange') + 
  geom_smooth()
```

```{r echo=FALSE, bivariate_alcohol6}
#trimming big outliers
ggplot(aes(x = alcohol, y = quality), data = df) + 
  geom_jitter(alpha = 0.1, color = 'orange') + 
  xlim(quantile(df$alcohol, 0.01), quantile(df$alcohol,0.99)) +
  geom_smooth()
```

## sulphates & quality

```{r echo=FALSE, , bivariate_sulphate1}
#plot the sulphates vs quality
ggplot(aes(quality, sulphates), data = df) +
  geom_jitter(alpha = 0.1)

```

```{r echo=FALSE, bivariate_sulphate2}
ggplot(aes(x= factor(quality), y = sulphates ), data = df) +
geom_jitter( alpha = .1)  +
  geom_boxplot( alpha = .3,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
```

```{r echo=FALSE, bivariate_sulphate3}
ggplot(aes(x = quality, y =sulphates), data = df) + 
  geom_boxplot(aes(group = quality))
```

```{r echo=FALSE, bivariate_sulphate4}
#removing big outliers
ggplot(aes(x = quality, y = sulphates), data = df) + 
  geom_boxplot(aes(group = quality)) + 
  ylim(quantile(df$sulphates, 0.01), quantile(df$sulphates, 0.99))
```

```{r echo=FALSE, bivariate_sulphate5}
ggplot(aes(x = sulphates, y = quality), data = df) + 
  geom_jitter(alpha = 0.1,  color = 'orange')  + 
  xlim(quantile(df$sulphates, 0.01), quantile(df$sulphates, 0.99)) + 
  geom_smooth()
```

## alcohol & sulphates

```{r echo=FALSE, bivariate_sulphate_alcohol}
with(subset(df, df$alcohol>11), cor.test(sulphates, 
                                         alcohol, method = 'pearson'))
```

```{r echo=FALSE, bivariate_sulphate_alcohol2}
ggplot(aes(alcohol, sulphates), data = df) +
  geom_point(position = 'jitter', alpha= 0.1)
```

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

```{r echo=FALSE, multivariate_alcohol_sulphate}
ggplot(aes(x = alcohol, 
           y = sulphates, color = factor(quality)), 
       data = df) + 
  ylim(quantile(df$sulphates, 0.01), quantile(df$sulphates, 0.99)) +
      geom_point(alpha = 0.8, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))
```

```{r echo=FALSE}
#Change the quality variable to an ordered factor
df$quality.int <- df$quality
df$quality <- factor(df$quality, ordered = TRUE)
str(df$quality)

```

```{r echo=FALSE}
ggplot(aes(density, alcohol, color = quality), data = df) +
  geom_point() +
  scale_color_brewer(type = 'qual')

```


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE}
ggplot(aes(density, alcohol, color = quality), data = df) +
  geom_point() +
  scale_color_brewer(type = 'qual')+
  ggtitle('Relationship between alchohol, density, and quality ')

```

### Description One
Alchohol has a positive relationship (.43) with quality. Generally, higher quality wines will have higher alchohol content. Alchohol has a negative relationship with density.

### Plot Two
```{r echo=FALSE}
ggplot(aes(x= factor(quality), y = alcohol ), data = df) +
geom_jitter( alpha = .1)  +
  geom_boxplot( alpha = .3,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  ggtitle('Relationship & distribution of alcohol and quality')
```

### Description Two

There's a high concentration of 5-7 quality wines... and these largely 
have alchohol levels between 8.5-13.5 

### Plot Three
```{r, echo=FALSE}

s0 = plot.hist('sulphates', df, 30)
s1 =  ggplot(df, aes(y = sulphates, x = 1)) + 
  geom_jitter(aes(x = 1, y = sulphates), alpha = 0.15) +
  geom_boxplot(color = 'blue', alpha = 1/4) + coord_flip()
s2 = plot.hist('sulphates', df, 75) 


title1="A deeper look at sulphates"

grid.arrange(s0, s1, s2+theme(legend.position="none"), top=title1,
             layout_matrix=matrix(c(1,1,2,3,4,4), ncol=2, byrow=TRUE),heights=c(1,1.5,0.3))


```


### Description Three
Sulphates range from .2 to 1.2 and are mostly centered around .5

------

# Reflection

EDA results: We reviewed three features in this data set: alcohol, quality, 
and sulphates. We found that alchohol has a positive relationship (.43) with 
quality. Generally, higher quality wines will have higher alchohol content. 
Sulphates and alchohol content generally do not have a strong relationship. 

Process reflection: Here we used plotting (scatterplots, histograms, 
boxplots, and line graphs). 

In the future, I would like to do a correlation matrix using all of the variables. 
I would also like to review ratios such as quality / alcohol. Secondly, I would 
like to do a deeper dive into alcohol and density.  Lastly, I would like to create 
a model which with observations can predict wine quality with some degree of confidence. 
And overall, I would like more practice with transparency, jitter, smoothing, and 
limiting axes. 
